Concentration of Maxima and Fundamental Limits in High-Dimensional Testing and Inference

### 2021-03-10 7.GWAS/7.odds-and-power.tex
ch 7
i implemented my minor edits to ch 7.  I have 3 paragraphs in BLUE which need to be checked -- pages 97, 102, 105.

In order to do so, we must first connect the concept of ``statistical signal size'' $\lambda$ with some key quantities in association tests.
While ``signal size'' likely sounds foreign to most practitioners, it is intimately linked with the concept of ``effect sizes'' --- or odds ratios --- in association studies, which are frequently estimated and reported in GWAS catalogs.
We characterize the relationship between the two quantities in the special, but fairly common case of association tests on 2-by-2 contingency tables in Section \ref{sec:odds-and-power}.

@@

Consider a 2-by-2 multinomial distribution with marginal probabilities of phenotypes $(\phi_1, \phi_2)$ and genotypes $(\theta_1, \theta_2)$.
The \emph{probability} table (as opposed to the table of multinomial \emph{counts} in the introduction) is as follows.
\begin{center}
    \begin{tabular}{cccc}
    \hline
    & \multicolumn{2}{c}{Genotype} \\
    \cline{2-3}
    Probabilities & Variant 1 & Variant 2 & Total by phenotype \\
    \hline
    Cases & $\mu_{11}$ & $\mu_{12}$ & $\phi_1$ \\
    Controls & $\mu_{21}$ & $\mu_{22}$ & $\phi_2$ \\
    Total by genotype & $\theta_1$ & $\theta_2$ & 1 \\
    \hline
    \end{tabular}
\end{center}
The odds ratio (i.e., ``effect size'') is defined as the ratio of the phenotype frequencies between the two genotype variants,
\begin{equation} \label{eq:odds-ratio}
    \text{R} := \frac{\mu_{11}}{\mu_{21}}\Big/\frac{\mu_{12}}{\mu_{22}}
    = \frac{\mu_{11}\mu_{22}}{\mu_{12}\mu_{21}}.
\end{equation}
The multinomial distribution is fully parametrized by the trio $(\theta_1, \phi_1, R)$.
Odds ratios further away from 1 indicate greater contrasts between the probability of outcomes.
Independence between the genotypes and phenotypes would imply an odds ratio of one, and hence $\mu_{jk} = \phi_j\theta_k$, for all $j,k \in\{1,2\}$.

@@

\begin{proof}
	We parametrize the 2-by-2 multinomial distribution with the parameter $\delta$, 
	\begin{equation} \label{eq:reparametrize-2-by-2-table-1}
	\mu_{11} = \phi_1\theta_1+\delta,\quad 
	\mu_{12} = \phi_1\theta_2-\delta,\quad 
	\mu_{21} = \phi_2\theta_1-\delta,\quad 
	\mu_{22} = \phi_2\theta_2+\delta.
	\end{equation}
	By relabelling of categories, we may assume $0<\theta_1,\phi_1\le1/2$ without loss of generality.
	Note that $\delta$ must lie within the range $[\delta_\mathrm{min}, \delta_\mathrm{max}]$, where
	$$
	\delta_\mathrm{min} := \max\{-\phi_1\theta_1, -\phi_2\theta_2, \phi_1\theta_2-1, \phi_2\theta_1-1\} 
	= -\phi_1\theta_1,
	$$
	and
	$$
	\delta_\mathrm{max} := \min\{1-\phi_1\theta_1, 1-\phi_2\theta_2, \phi_1\theta_2, \phi_2\theta_1\}
	= \min\{\phi_1\theta_2, \phi_2\theta_1\},
	$$
	in order for $\mu_{ij}\ge0$ for all $i,j\in \{1,2\}$.
	Under this parametrization, Relation \eqref{eq:odds-ratio} then becomes
	\begin{equation} \label{eq:odds-ratio-delta}
	\text{R} = \frac{\mu_{11}\mu_{22}}{\mu_{12}\mu_{21}}
	= \frac{\phi_1\theta_1\phi_2\theta_2 + \delta(\phi_1\theta_1+\phi_2\theta_2)+\delta^2}{\phi_1\theta_1\phi_2\theta_2 - \delta(\phi_1\theta_2+\phi_2\theta_1)+\delta^2},
	\end{equation}
	which is one-to-one and increasing in $\delta$ on $(\delta_\mathrm{min}, \delta_\mathrm{max})$.
	Equation \eqref{eq:signal-size-chisq} becomes
	\begin{equation} \label{eq:signal-size-chisq-delta}
	w^2 = \sum_{i=1}^2 \sum_{j=1}^2 \frac{(\mu_{ij} - \phi_i\theta_j)^2}{\phi_i\theta_j}
	= \delta^2\sum_i\sum_j \frac{1}{\phi_i\theta_j}
	= \frac{\delta^2}{\phi_1\theta_1\phi_2\theta_2},
	\end{equation}
	Solving for $\delta$ in \eqref{eq:odds-ratio-delta}, and plugging into the expression for signal size \eqref{eq:signal-size-chisq-delta} yields Relation \eqref{eq:signal-size-odds-ratio}.

>>>>

In order to do so, we must first connect the concept of   ``statistical signal size'' $\lambda$ with some key quantities in association tests.
While ``signal size'' likely sounds foreign to most practitioners, it is intimately linked with the concept of ``effect sizes'' --- or odds ratios --- in association studies, which are frequently estimated and reported in GWAS catalogs. This terminology, in turn, may be alien to some statisticians.  Here, we aim to clarify 
the two languages.

 %in Section \ref{sec:odds-and-power}.

@@

\stilian{Recall the general setup of genetic association testing in Chapter \ref{sec:motivation-chisq}, where one wants to test 
for the effect (or lack thereof) of a specific gene on the occurrence of a disease.  A randomly drawn individual from the entire population 
is characterized by a pair $(P,G)$ of binary random variables, where $P$ encodes the phenotype and $G$, the genotype.  For example, in
the context of a specific study,  by phenotype we mean whether or not the individual has the disease (belongs to the Cases group) or is healthy 
(Controls group).  The genotype encodes whether or not the gene in question is expressed (Variant 1) or not (Variant 2). 
Table  \ref{tab:multinomial-counts} in the introduction summarizes the {\em counts} for all phenotype-genotype combinations
for the individuals in a given study sampled from the population. These counts can be assumed to follow a multinomial distribution, with probabilities 
given in the following Table \ref{tab:multinomial-probabilites}.  Notice that there are fundamental study design questions here since the counts will 
depend on the way individuals are recruited in the study.  For example, ``controlling'' the number of Cases or Controls (as opposed to sampling at
random from the population) will change the theoretical parameters (probabilities) of the multinomial model.  
These questions will be addressed in Section \ref{sec:optimal-design}  below.

In this section, we characterize the relationship between the aforementioned ``signal size'' and ``odds-ratio'' parameterizations 
in the special, but fairly common case of association tests on 2-by-2 contingency tables.
}{\fbox{CHECK PLEASE}}
Consider a 2-by-2 multinomial distribution with marginal probabilities of phenotypes $(\phi_1, \phi_2)$ and genotypes $(\theta_1, \theta_2)$.
The \emph{probability} table (as opposed to the table of multinomial \emph{counts} in the introduction) is as follows.
\begin{table}[h]\label{tab:multinomial-probabilites}
\begin{center}
    \begin{tabular}{cccc}
    \hline
    & \multicolumn{2}{c}{Genotype} \\
    \cline{2-3}
    Probabilities & Variant 1 & Variant 2 & Total by phenotype \\
    \hline
    Cases & $\mu_{11}$ & $\mu_{12}$ & $\phi_1$ \\
    Controls & $\mu_{21}$ & $\mu_{22}$ & $\phi_2$ \\
    Total by genotype & $\theta_1$ & $\theta_2$ & 1 \\
    \hline
    \end{tabular}
    \caption{Probabilities of the multinomial distribution in a genetic association test. (Compare and contrast with 
    Table \ref{tab:multinomial-counts}. We have $\E[ O_{ij}] = n\mu_{ij},\ i,j=1,2$, where $n = \sum_{i,j} O_{ij}$.) }
\end{center}
\end{table}

The odds ratio (i.e., ``effect size'') is defined as the ratio of the phenotype frequencies between the two genotype variants,
\begin{equation} \label{eq:odds-ratio}
    \text{R} := \frac{\mu_{11}}{\mu_{21}}\Big/\frac{\mu_{12}}{\mu_{22}}
    = \frac{\mu_{11}\mu_{22}}{\mu_{12}\mu_{21}}.
\end{equation}
The multinomial distribution is fully parametrized by the trio $(\theta_1, \phi_1, R)$.
Odds ratios further away from 1 indicate greater contrasts between the probability of outcomes.
Independence between the genotypes and phenotypes would imply an odds ratio of one, and hence $\mu_{jk} = \phi_j\theta_k$, for all $j,k \in\{1,2\}$.

@@

\begin{proof}
	We parametrize the 2-by-2 multinomial distribution with the parameter $\delta$, 
	\begin{equation} \label{eq:reparametrize-2-by-2-table-1}
	\mu_{11} = \phi_1\theta_1+\delta,\quad 
	\mu_{12} = \phi_1\theta_2-\delta,\quad 
	\mu_{21} = \phi_2\theta_1-\delta,\quad 
	\mu_{22} = \phi_2\theta_2+\delta.
	\end{equation}
	By relabelling of categories, we may assume $0<\theta_1,\phi_1\le1/2$ without loss of generality.
	Note that $\delta$ must lie within the range $[\delta_\mathrm{min}, \delta_\mathrm{max}]$, where
	$$
	\delta_\mathrm{min} := \max\{-\phi_1\theta_1, -\phi_2\theta_2, \phi_1\theta_2-1, \phi_2\theta_1-1\} 
	= -\phi_1\theta_1,
	$$
	and
	$$
	\delta_\mathrm{max} := \min\{1-\phi_1\theta_1, 1-\phi_2\theta_2, \phi_1\theta_2, \phi_2\theta_1\}
	= \min\{\phi_1\theta_2, \phi_2\theta_1\},
	$$
	in order for $\mu_{ij}\ge0$ for all $i,j\in \{1,2\}$.
	Under this parametrization, Relation \eqref{eq:odds-ratio} then becomes
	\begin{align} \label{eq:odds-ratio-delta}
	\text{R} = \frac{\mu_{11}\mu_{22}}{\mu_{12}\mu_{21}}
	= \frac{\phi_1\theta_1\phi_2\theta_2 + \delta(\phi_1\theta_1+\phi_2\theta_2)+\delta^2}{\phi_1\theta_1\phi_2\theta_2 - \delta(\phi_1\theta_2+\phi_2\theta_1)+\delta^2}
	= \frac{A +\delta B + \delta^2}{A -\delta C + \delta^2},
	\end{align}
	which is one-to-one and increasing in $\delta$ on $(\delta_\mathrm{min}, \delta_\mathrm{max})$.
	Equation \eqref{eq:signal-size-chisq} becomes
	\begin{equation} \label{eq:signal-size-chisq-delta}
	w^2 = \sum_{i=1}^2 \sum_{j=1}^2 \frac{(\mu_{ij} - \phi_i\theta_j)^2}{\phi_i\theta_j}
	= \delta^2\sum_i\sum_j \frac{1}{\phi_i\theta_j}
	= \frac{\delta^2}{\phi_1\theta_1\phi_2\theta_2},
	\end{equation}
	Solving for $\delta$ in \eqref{eq:odds-ratio-delta}, and plugging into the expression for signal size \eqref{eq:signal-size-chisq-delta} yields Relation \eqref{eq:signal-size-odds-ratio}.

### 2021-03-10 7.GWAS/7.optimal-design.tex
ch 7
i implemented my minor edits to ch 7.  I have 3 paragraphs in BLUE which need to be checked -- pages 97, 102, 105.

Fortunately, the conditional distributions of genotypes in the healthy control groups are often estimated by existing studies, and are made available by consortia such as the NHGRI-EBI GWAS catalog \citep{macarthur2016new}.
% Assume (after appropriate relabelling, hence without loss of generality) that the first variant is associated with an increased risk of disease, and is henceforth referred to as the risk variant.
We denote the conditional frequency of the first genetic variant in the control group as $(f, 1-f)$, where
\begin{equation} \label{eq:RAF}
    f := \mu_{21} / \phi_2.
\end{equation}
The multinomial probability is fully parametrized by the new trio: $(f, \phi_1, R)$.
\begin{center}
    \begin{tabular}{cccc}
    \hline
    & \multicolumn{2}{c}{Genotype} \\
    \cline{2-3}
    Probabilities & Variant 1 & Variant 2 & Total by phenotype \\
    \hline
    Cases & $\frac{\phi_1fR}{fR+1-f}$ & $\frac{\phi_1(1-f)}{fR+1-f}$ & $\phi_1$ \\
    Controls & $f(1-\phi_1)$ & $(1-f)(1-\phi_1)$ & $1-\phi_1$ \\
    \hline
    \end{tabular}
\end{center}
Proposition \ref{prop:signal-size-odds-ratio} may also be re-stated in terms of the new parametrization.

@@

\begin{proof}
	Using the parametrization in \eqref{eq:reparametrize-2-by-2-table-1}, we solve for $\delta$ in \eqref{eq:odds-ratio-delta} to obtain
	\begin{align}
		\delta &= \frac{\phi_1 fR}{fR+1-f} - \left(\frac{\phi_1 fR}{fR+1-f} + f(1-\phi_1)\right)\phi_1 \nonumber \\
		&= \frac{f(1-f)\phi_1(1-\phi_1)(R-1)}{fR+1-f}. \label{eq:reparametrize-2-by-2-table-2}
	\end{align}
	Substituting \eqref{eq:reparametrize-2-by-2-table-2} into the expression \eqref{eq:signal-size-chisq-delta}, after some simplification, yields
	\begin{equation} \label{eq:reparametrize-2-by-2-table-3}
	w^2 = \frac{f(1-f)\phi_1(1-\phi_1)(R-1)^2}{\left[\phi_1 R + (1-\phi_1)D\right]\left[\phi_1 + (1-\phi_1)D\right]},
	\end{equation}
	where $D = fR+1-f > 0$.
	Therefore, he derivative of \eqref{eq:reparametrize-2-by-2-table-3} with respect to $\phi_1$ is
	\begin{equation} \label{eq:signal-size-first-derivative}
	\frac{\mathrm{d}w^2}{\mathrm{d}\phi_1} = 
	\frac{f(1-f)(R-1)^2}{\left[\phi_1 R+(1-\phi_1)D\right]^2 \left[\phi_1+(1-\phi_1)D\right]^2} \left[(D^2-R)\phi_1^2 - 2D^2\phi_1 + D^2\right].
	\end{equation}
	Further, we obtain the second derivative with respect to $\phi_1$,
	\begin{equation} \label{eq:signal-size-second-derivative}
	\frac{\mathrm{d}^2w^2}{\mathrm{d}\phi_1^2} = 
	h(R,f) \left[(\phi_1-1)D^2 - \phi_1R\right],
	\end{equation}
	where $h$ is some function of $(R,f)$ taking on strictly positive values.

	Since $\left[(\phi_1-1)D^2 - \phi_1R\right]<0$, the second derivative \eqref{eq:signal-size-second-derivative} must be strictly negative on $[0,1]$.
	This implies that the first derivative \eqref{eq:signal-size-first-derivative} is strictly decreasing on $[0,1]$. 
	Since the first derivative \eqref{eq:signal-size-first-derivative} is strictly positive at $\phi_1=0$, and strictly negative at $\phi_1=1$, it must have a unique zero between 0 and 1, and hence, the solution to $(D^2-R)\phi_1^2 - 2D^2\phi_1 + D^2 = 0$ in the interval of $[0,1]$ must be the maximizer of \eqref{eq:reparametrize-2-by-2-table-3} --- when $D^2-R>0$, the smaller of the two roots maximizes \eqref{eq:reparametrize-2-by-2-table-3}, and when $D^2-R<0$, it is the larger of the two.
	They share the same expression ${D}/{(D+\sqrt{R})}$, which coincides with \eqref{eq:optimal-design}.
	Finally, when $D^2=R$, the only root $\phi_1^*=1/2$, which also coincides with \eqref{eq:optimal-design}, is the maximizer of \eqref{eq:reparametrize-2-by-2-table-3}.
\end{proof}

>>>>

Fortunately, the conditional distributions of genotypes in the healthy control groups are often estimated by existing studies, and are made available by consortia such as the NHGRI-EBI GWAS catalog \citep{macarthur2016new}.
% Assume (after appropriate relabelling, hence without loss of generality) that the first variant is associated with an increased risk of disease, and is henceforth referred to as the risk variant.
We denote the conditional frequency of the first genetic variant in the control group as $(f, 1-f)$, where
\begin{equation} \label{eq:RAF}
    f := \mu_{21} / \phi_2 = \mu_{21}/ (1-\phi_1).
\end{equation}
The multinomial probability is fully parametrized by the new trio: $(f, \phi_1, R)$.
\begin{center}
    \begin{tabular}{cccc}
    \hline
    & \multicolumn{2}{c}{Genotype} \\
    \cline{2-3}
    Probabilities & Variant 1 & Variant 2 & Total by phenotype \\
    \hline
    Cases & $\frac{\phi_1fR}{fR+1-f}$ & $\frac{\phi_1(1-f)}{fR+1-f}$ & $\phi_1$ \\
    Controls & $f(1-\phi_1)$ & $(1-f)(1-\phi_1)$ & $1-\phi_1$ \\
    \hline
    \end{tabular}
\end{center}
Proposition \ref{prop:signal-size-odds-ratio} may also be re-stated in terms of the new parametrization.

@@

\begin{proof}
	Using the parametrization in \eqref{eq:reparametrize-2-by-2-table-1}, we solve for $\delta$ in \eqref{eq:odds-ratio-delta} to obtain
	\begin{align}
		\delta &= \frac{\phi_1 fR}{fR+1-f} - \left(\frac{\phi_1 fR}{fR+1-f} + f(1-\phi_1)\right)\phi_1 \nonumber \\
		&= \frac{f(1-f)\phi_1(1-\phi_1)(R-1)}{fR+1-f}. \label{eq:reparametrize-2-by-2-table-2}
	\end{align}
	Substituting \eqref{eq:reparametrize-2-by-2-table-2} into the expression \eqref{eq:signal-size-chisq-delta}, after some simplification, yields
	\begin{equation} \label{eq:reparametrize-2-by-2-table-3}
	w^2 = \frac{f(1-f)\phi_1(1-\phi_1)(R-1)^2}{\left[\phi_1 R + (1-\phi_1)D\right]\left[\phi_1 + (1-\phi_1)D\right]},
	\end{equation}
	where $D = fR+1-f > 0$.
	Therefore, the derivative of \eqref{eq:reparametrize-2-by-2-table-3} with respect to $\phi_1$ is
	\begin{equation} \label{eq:signal-size-first-derivative}
	\frac{\mathrm{d}w^2}{\mathrm{d}\phi_1} = 
	\frac{f(1-f)(R-1)^2}{\left[\phi_1 R+(1-\phi_1)D\right]^2 \left[\phi_1+(1-\phi_1)D\right]^2} \left[(D^2-R)\phi_1^2 - 2D^2\phi_1 + D^2\right].
	\end{equation}
	Further, we obtain the second derivative with respect to $\phi_1$,
	\begin{equation} \label{eq:signal-size-second-derivative}
	\frac{\mathrm{d}^2w^2}{\mathrm{d}\phi_1^2} = 
	h(R,f) \left[(\phi_1-1)D^2 - \phi_1R\right],
	\end{equation}
	where $h$ is some function of $(R,f)$ taking on strictly positive values.

	Since $\left[(\phi_1-1)D^2 - \phi_1R\right]<0$, the second derivative \eqref{eq:signal-size-second-derivative} must be strictly negative on $[0,1]$.
	This implies that the first derivative \eqref{eq:signal-size-first-derivative} is strictly decreasing on $[0,1]$. 
	Since the first derivative \eqref{eq:signal-size-first-derivative} is strictly positive at $\phi_1=0$, and strictly negative at $\phi_1=1$, it must have a unique zero between 0 and 1, and hence, the solution to $(D^2-R)\phi_1^2 - 2D^2\phi_1 + D^2 = 0$ in the interval of $[0,1]$ must be the maximizer of \eqref{eq:reparametrize-2-by-2-table-3} --- when $D^2-R>0$, the smaller of the two roots maximizes \eqref{eq:reparametrize-2-by-2-table-3}, and when $D^2-R<0$, it is the larger of the two.
	They share the same expression ${D}/{(D+\sqrt{R})}$, which coincides with \eqref{eq:optimal-design}.
	Finally, when $D^2=R$, the only root $\phi_1^*=1/2$, which also coincides with \eqref{eq:optimal-design}, is the maximizer 
	of \eqref{eq:reparametrize-2-by-2-table-3}.  \stilian{This is so, in particular, when 
	$R=1$ and confirms our intuition that the symmetric design is optimal when the odds ratio 
	equals one.  This is not the only case when symmetrical designs are optimal though. }{\fbox{ Am I CORRECT?}}
\end{proof}

### 2021-03-10 7.GWAS/GWAS.tex
rename files and folders according to the new structure
ch 7
i implemented my minor edits to ch 7.  I have 3 paragraphs in BLUE which need to be checked -- pages 97, 102, 105.

The process of scientific discovery, as famously explained by Richard Faynman, usually starts with guesses in one form or another.
The consequences of such guesses are then computed, and compared with experimental results.
If the predictions disagree with experiment, then our guesses are wrong. ``That is all there is to it''  \citep{feynman2017character}.

In previous chapters, we delved deep in to the theoretical underpinnings of the phase transition phenomena in high-dimensional multiple testing problems.
The results are interesting in their own right.
However, in the spirit of Faynman, we have not discovered any scientific law, but merely worked out mathematical consequences of our postulated models.
In this chapter, our goal is to relate these predictions to real experimental data from the field of genetics, where large-scale simultaneous hypotheses testing problems often arise.
From such comparisons, we will demonstrate that the phase transition laws are indeed accurate predictions of some curious phenomena in the real world.
The accuracy of our predictions will lend credibility to the application of these ``laws of large dimensions'' in actual statistical problems.

@@

In our case, the experimental data used as the measuring stick come from genome-wide association studies (\ac{GWAS}), introduced in Section \ref{sec:motivation-chisq}.
Recall that in \ac{GWAS}, a large number of marginal association tests are conducted simultaneously, resulting in statistics that can be approximated by
\begin{equation} \label{eq:model-chisq-Chapter6}
    %x(i) \distras{\mathrm{ind.}} \chi_\nu^2\left(\lambda(i)\right), \quad i=1,\ldots,p.
    x(i) \sim \chi_\nu^2\left(\lambda(i)\right), \quad i=1,\ldots,p,
\end{equation}
where $\chi_\nu^2\left(\lambda(i)\right)$ is a chi-square distributed random variable with $\nu$ degrees of freedom and non-centrality parameter $\lambda(i)$.

We establish our theoretical predictions in two steps.
In Section \ref{sec:chisq-boundaries} below, we shall first establish the phase transitions of the model \eqref{eq:model-chisq-Chapter6}.
In parallel to results in Chapter \ref{chap:phase-transitions}, we show that several commonly used family-wise error rate-control procedures --- including Bonferroni's procedure --- are asymptotically optimal for the {exact}, and {exact-approximate} support recovery problems (recall Definition \ref{def:exact-recovery-success-failure}) in idealized chi-square models with independent components.
Analogously, the \ac{BH} procedure is asymptotically optimal for the {approximate}, and {approximate-exact} support recovery problems.
% \ref{thm:chi-squared-exact-boundary}, \ref{thm:chi-squared-exact-approx-boundary}, \ref{thm:chi-squared-approx-boundary}, and \ref{thm:chi-squared-approx-exact-boundary}.
Under appropriate parametrizations of the signal sizes and sparsity, they establish the phase transitions of support recovery problems in the chi-square model.
Remarkably, the degree-of-freedom parameter does not affect the asymptotic boundaries in any of the four support recovery problems.

We then translate the canonical signal size and sparsity parametrizations into the vernacular of statistical geneticists in Section \ref{sec:odds-and-power}.
% present empirical evidence for the phase transition in the exact-approximate problem using real data from large-scale association studies on breast cancer obtained from the NHGRI-EBI GWAS Catalog \citep{macarthur2016new}.
% demystify the notion of signal size $\lambda$ in this context.
% which is perhaps less transparent than in additive error models.
We do so by characterizing the relationship between the signal size $\lambda$ and the marginal frequencies, odds ratio, and sample sizes for association tests on 2-by-2 contingency tables. 
These parameters are more often estimated and reported in \ac{GWAS} than the elusive signal size parameter $\lambda$.
% Specifically, the amount of signal, when rare variants are present, is weaker compared to the signal when the marginal distributions are balanced.
% In other words, reliable detection of the effects by rare variants would require more samples compared to common variants, even at the same odds ratio.
% This result, establishing the relationship between sample sizes and signal sizes, is made precise in Section \ref{sec:odds-and-power}.
As a bonus, we point out an implications of this relationship on optimal study designs for association studies in Section \ref{sec:optimal-design}.
Perhaps surprisingly, our analysis reveals that balanced designs with equal number of cases and controls are often statistically inefficient.
% Practical consequences of these results in power analysis will be illustrated with data examples in Section \ref{sec:phase-transitions-in-GWAS}.

Armed with the results on phase transitions in the chi-square model, and a translation from the language of high-dimenstional statistics to the patois of association screening studies, we finally present in Section \ref{sec:phase-transitions-in-GWAS} the consequences of the phase transitions in \ac{GWAS}, and compare against real experimental data to evaluate the success of our predictions.

The phase transitions in the chi-square models are demonstrated with numerical simulations in Section \ref{sec:numerical}.
Proofs are collected in the online Supplement.
%in Section \ref{sec:proof-signal-size-odds-ratio}.

@@

\section{Support recovery problems in chi-squared models}
\label{sec:chisq-boundaries}
\input{6.GWAS/6.chi-square.tex}

\section{Odds ratios and statistical power}
\label{sec:odds-and-power}
\input{6.GWAS/6.odds-and-power.tex}

\section{Optimal study designs and rare variants}
\label{sec:optimal-design} 
\input{6.GWAS/6.optimal-design.tex}

\section{Phase transitions in large-scale association screening studies}
\label{sec:phase-transitions-in-GWAS}
\input{6.GWAS/6.GWAS-phase-transitions.tex}

\section{Numerical illustrations of the phase transitions in chi-square models}
\label{sec:numerical}
\input{6.GWAS/6.numerical.tex}

>>>>

The process of scientific discovery, as famously explained by Richard Feynman, usually starts with guesses in one form or another.
The consequences of such guesses are then computed, and compared with experimental results.
If the predictions disagree with the experiment, then our guesses are wrong. ``That is all there is to it''  \citep{feynman2017character}.

In the previous chapters, we delved deep into the theoretical underpinnings of the phase transition phenomena in high-dimensional multiple testing problems.
The results are interesting in their own right.
However, in the spirit of Feynman, we have not discovered any scientific law, but merely worked out mathematical consequences of our postulated models.
In this chapter, our goal is to relate these predictions to real experimental data from the field of genetics, where large-scale simultaneous hypotheses testing problems often arise.
From such comparisons, we will demonstrate that the phase transition laws are indeed accurate predictions of some curious phenomena in the real world.
The accuracy of our predictions will lend credibility to the application of these ``laws of large dimensions'' in actual statistical problems.

@@

In our case, the experimental data used as the measuring stick come from genome-wide association studies (\ac{GWAS}), introduced in Section \ref{sec:motivation-chisq}.
Recall that in \ac{GWAS}, a large number of marginal association tests are conducted simultaneously, resulting in statistics that can be approximated by
\begin{equation} \label{eq:model-chisq-Chapter6}
    %x(i) \distras{\mathrm{ind.}} \chi_\nu^2\left(\lambda(i)\right), \quad i=1,\ldots,p.
    x(i) \sim \chi_\nu^2\left(\lambda(i)\right), \quad i=1,\ldots,p,
\end{equation}
where $\chi_\nu^2\left(\lambda(i)\right)$ is a chi-square distributed random variable with $\nu>0$ degrees of 
freedom\footnote{The parameter $\nu$ here should not be confused with the shape parameter of the AGG$(\nu)$ 
distributions, which will not appear in this chapter.} and non-centrality parameter $\lambda(i)$.

We establish our theoretical predictions in two steps.
In Section \ref{sec:chisq-boundaries} below, we shall first establish the phase transitions of the model \eqref{eq:model-chisq-Chapter6}.
In parallel to results in Chapter \ref{chap:phase-transitions}, we show that several commonly used family-wise error rate-control procedures --- including Bonferroni's procedure --- are asymptotically optimal for the {exact}, and {exact-approximate} support recovery problems (recall Definition \ref{def:exact-recovery-success-failure}) in idealized chi-square models with independent components.
Analogously, the \ac{BH} procedure is asymptotically optimal for the {approximate}, and {approximate-exact} support recovery problems.
% \ref{thm:chi-squared-exact-boundary}, \ref{thm:chi-squared-exact-approx-boundary}, \ref{thm:chi-squared-approx-boundary}, and \ref{thm:chi-squared-approx-exact-boundary}.
Under appropriate parametrization of the signal sizes and sparsity, we establish the phase transitions of support recovery problems in the chi-square model.
Remarkably, the degree-of-freedom parameter does not affect the asymptotic boundaries in any of the four support recovery problems.

We then translate the canonical signal size and sparsity parametrizations into the vernacular of statistical geneticists in Section \ref{sec:odds-and-power}.
% present empirical evidence for the phase transition in the exact-approximate problem using real data from large-scale association studies on breast cancer obtained from the NHGRI-EBI GWAS Catalog \citep{macarthur2016new}.
% demystify the notion of signal size $\lambda$ in this context.
% which is perhaps less transparent than in additive error models.
We do so by characterizing the relationship between the signal size $\lambda$ and the marginal frequencies, odds ratio, and sample sizes for association tests on 2-by-2 contingency tables. 
These parameters are more often estimated and reported in \ac{GWAS} than the elusive signal size parameter $\lambda$.
% Specifically, the amount of signal, when rare variants are present, is weaker compared to the signal when the marginal distributions are balanced.
% In other words, reliable detection of the effects by rare variants would require more samples compared to common variants, even at the same odds ratio.
% This result, establishing the relationship between sample sizes and signal sizes, is made precise in Section \ref{sec:odds-and-power}.
As a bonus, we point out the implications of this relationship on statistically optimal study designs for association studies in Section \ref{sec:optimal-design}.
Perhaps surprisingly, our analysis reveals that balanced designs with equal number of cases and controls are often statistically inefficient.
% Practical consequences of these results in power analysis will be illustrated with data examples in Section \ref{sec:phase-transitions-in-GWAS}.

Armed with the results on phase transitions in the chi-square model, and a translation from the language of high-dimenstional statistics to the {\em patois} of association screening studies, we finally present in Section \ref{sec:phase-transitions-in-GWAS} the consequences of the phase transitions in \ac{GWAS}, and compare against real experimental data to evaluate the success of our predictions.

The phase transitions in the chi-square models are demonstrated with numerical simulations in Section \ref{sec:numerical}.
The proofs, which are closely related to those of the results in Chapter \ref{chap:phase-transitions},
 are collected in Section \ref{sec:supplement:proofs}.
%in Section \ref{sec:proof-signal-size-odds-ratio}.

@@

\section{Support recovery problems in chi-squared models}
\label{sec:chisq-boundaries}
\input{7.GWAS/7.chi-square.tex}

\section{Odds ratios and statistical power}
\label{sec:odds-and-power}
\input{7.GWAS/7.odds-and-power.tex}

\section{Optimal study designs and rare variants}
\label{sec:optimal-design} 
\input{7.GWAS/7.optimal-design.tex}

\section{Phase transitions in large-scale association screening studies}
\label{sec:phase-transitions-in-GWAS}
\input{7.GWAS/7.GWAS-phase-transitions.tex}

\section{Numerical illustrations of the phase transitions in chi-square models}
\label{sec:numerical}
\input{7.GWAS/7.numerical.tex}
